---
title: "formulas"
format: gfm
editor: visual
execute: 
  echo: false
  warning: false
---

```{r}
library(tidyverse)
library(yingtools2)
library(knitr)
```

Various distance metrics have been used to assess beta-diversity across sample populations in microbial ecology.

Some are shown below, where $D_{jk}$ denotes the calculated distance between samples $j$ and $k$, where $x_{ij}$ and $x_{ik}$ refer to the quantity of bacterial species/strain $i$ in pairwise samples $j$ and $k$.

```{r}
values <- c("Euclidean"="$\\textrm{Deuc}_{jk} = \\sqrt{\\sum_i (x_{ij}-x_{ik})^2}$",
            "Manhattan"="$\\textrm{Dman}_{jk}=\\sum_i \\mid x_{ij}-x_{ik} \\mid$",
            "Morisita"="$\\textrm{Dmor}_{jk} =  1 - \\dfrac{2 \\sum_i x_{ij} x_{ik}}{(\\lambda_j +
              \\lambda_k) \\sum_i x_{ij} \\sum_i
              x_{ik}}$, where $\\lambda_j = \\dfrac{\\sum_i x_{ij} (x_{ij} - 1)}{\\sum_i x_{ij} \\sum_i (x_{ij} - 1)}$",
            "Morisita-Horn"="$\\textrm{Dhorn}_{jk} =  1 - \\dfrac{2 \\sum_i x_{ij} x_{ik}}{(\\lambda_j +
              \\lambda_k) \\sum_i x_{ij} \\sum_i
              x_{ik}}$, where $\\lambda_j = \\sum_i {x_{ij}}^2/(\\sum_i x_{ij})^2$")
tbl <- tibble("Distance Metric"=names(values),'Definition'=values)

kable(tbl)
```

Other metrics calculate beta diversity by examining genetic related-ness among microbial members. Unifrac calculates beta-diversity distance between two populations using a phylogenetic tree, where $l_n$ represents the length between node $n$ and its parent, and $X_{nj}$ and $X_{nk}$ are indicators (0 or 1) as descendants of node $n$ are absent or present in samples $j$ and $k$, respectively.

```{r}
values2 <- c("Unweighted Unifrac"="$\\textrm{Dunifrac}_{jk} = \\dfrac{\\sum_n l_n \\mid A_n - B_n \\mid}{\\sum_n l_n \\max(X_{nj}, X_{nk})}$")
tbl2 <- tibble("Distance Metric"=names(values2),'Definition'=values2)

kable(tbl2)
```

There are limitations with all of the above metrics.



We used Horn index to calculate $\textrm{Dhorn}_{jk}$, representing the Horn calculation across all taxonomic levels

Let $H_{ijk}$ represent the Horn index calculated at the taxonomic level $i$ for samples $j$ and $k$.

$$
\textrm{Dtaxhorn}_{jk} = \dfrac{\sum_{i=2}^{N} (N-i-1) \textrm{Dhorn}_{ijk}}{\sum_{i=2}^{N} (N-i-1)}
$$

That is, the weighted average of Horn distances calculated for each taxonomic level. Lower taxonomic levels are weighted more than higher taxonomic levels. We omit the first level, Superkingdom, since its value is 'Bacteria' for all microbial members in the sample data. Therefore the weights across levels are as follows: Superkingdom: 0, Phylum: 7, Class: 6, Order: 5, Family: 4, Genus: 3, Species: 2, ASV: 1.



